Seasonality in governing upper-ocean dynamics at submesoscales

### 2016-06-09 writeup/rocha_etal.tex
Small fix in sec3
Significant improvement in the text

A key aspect of the LLC simulations is that there were forced by
the 16 most significant tidal components.
Because barotropic tides interact with topography and generate internal
tides that project on mesoscales to submesoscales
\citep[e.g., ][]{rocha_etal2016}, tidal forcing fundamentally distinguishes our analysis
from recent modeling studies that investigated upper-ocean dynamics
and its seasonality \citep{sasaki_etal2014,qiu_etal2014}. Details of the LLC simulations
are provided in the supplemental material.

@@

\begin{figure*}[ht]
\begin{center}
\hspace{-1.25cm}\includegraphics[width=.7\textwidth]{figs/fig1_1.pdf}\\
\vspace{-.125cm}
\includegraphics[width=.9\textwidth]{figs/fig1_2.pdf}
 \caption{(a) The study region with the subregion where the LLC outputs are
          analyzed. LLC 4320 (1/48$^\circ$) snapshots of surface vorticity (b and c) and a transect
          of potential density at 165$^\circ$E (d and e). The snapshots were
          taken at 00:00 UTC.}
\vspace{-1.5cm}
 \label{fig1}
 \end{center}
 \end{figure*}

@@

 \begin{figure*}[ht]
   \begin{center}
     \includegraphics[width=.8\textwidth]{figs/fig2.pdf}
  \caption{Time series of the root-mean-square (RMS) of vorticity (a) and
  lateral rate of strain (b), and horizontal divergence (c) in the LLC outputs and gridded AVISO data.}
  \label{fig2}
  \end{center}
\end{figure*}

@@

The seasonal cycle of the horizontal divergence, however, showcases the complexity
of the upper-ocean annual variability. If submesoscale eddies and fronts dominated
the near-surface variability all year, then the seasonal cycle of horizontal divergence,
vertical vorticity, and lateral strain rate would be in phase \citep[e.g.,][]{sasaki_etal2014}.
While there is a clear wintertime peak in divergence of daily-averaged velocity
(see dashed lines in figure \ref{fig2}a; RMS divergence $\sim$0.22 f in the 1/48$^\circ$
simulation), the hourly fields show
a stronger enhancement of
lateral divergence in late Summer/early Fall (RMS divergence $\sim$0.22 f the 1/48$^\circ$
simulation). Because the 1/48$^\circ$ simulation better resolve
smaller-scale submesoscale flows, the
winter peak in divergence is nearly as strong the summertime's. Of course, part of
these submesoscale fronts and eddies evolve relatively fast and there is no clear
temporal and spatial scale separation between those motions and inertia-gravity waves:
daily-averaging the velocity fields efficiently suppresses the summertime horizontally
divergent flows, but also reduces the wintertime lateral divergence by about 50$\%$.

\section{Seasonality in governing submesoscale dynamics}
The overall results of figure \ref{fig2}c strongly suggest that the
summertime submesoscale surface variability stems from different dynamics than
then wintertime variability. To better understand these differences, we calculate
joint probability distributions...

\begin{figure*}[ht]
  \begin{center}
    \includegraphics[width=.9\textwidth]{figs/fig3.pdf}
 \caption{Seasonal variation of joint probability distributions: vorticity vs.
          divergence (a through e) and vorticity vs. strain rate (f through j).
          Dashed lines in (a) through (e) represent $\delta = \frac{\omega}{f_0}
          \zeta$ where $\omega = 2\pi/$(12 h, 3 h) and $f_0$ is the inertial frequency
          in the middle of the domain (32.5$^\circ$ N). Dashed lines in (f) through (j)
          represent pure strain flow $\alpha = \zeta$.}
 \label{fig3}
 \end{center}
\end{figure*}

\section{Projection on horizontal scales}

To better understand the projection of these flows onto different horizontal
scales, we calculate wavenumber spectra of kinetic energy and sea-surface height
(SSH) variance. For simplicity we calculate zonal wavenumber spectra at various
latitudes. The two-year time averaged and spatial linear trend were
removed, and the ensuing fields were Hanning spectral ``window''. Statistical significance
is obtained by averaging over latitude and time. Using conservative temporal and
spatial decorrelation estimate of 10 days and 250 km, we obtain roughly 50
independent realizations of the spectra.

Figure \ref{fig3}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the spectra based on hourly
velocity snapshots are nearly indistinguishable within 95$\%$ confidence level
(compare solid lines in figure \ref{fig3}a). Consistent with the results of
\citet{rocha_etal2016} who analyzed the LLC4320 output in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales smaller than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig3}a). But this suppression is more dramatic in October, when the horizontally
divergent flows peak. At scales
smaller than 100 km, 40$\%$ of the surface KE in April is accounted for by supra-inertial
flows as opposed to 77$\%$ in October. The seasonality of subinertial submesoscale flows
is significant, consistent with the results of \citet{sasaki_etal2014}, which are
based on snapshots of a model without tidal forcing.

These high-frequency submesoscale flows significantly project on the sea-surface.
In October, when the horizontally divergent flows peak and the submesoscale
balanced flows are minimum, there is a dramatic difference between the spectra
based on hourly and daily-averaged SSH: at scales smaller than 100 km, the spectra
of hourly SSH roughly follows a $-2$ power-law, whereas the spectra of daily-averaged
SSH roughly follows a $-5$ power-law. At scales
smaller than 100 km, 44$\%$ of the surface KE in April is accounted for by suprainertial
flows as opposed to 86$\%$ in October! The out-of-phase seasonal cycle of subinertial
and suprainertial flows conspire to yield nearly insignificant seasonality in the spectra
of hourly SSH.

@@

\begin{figure*}[ht]
  \begin{center}
    \includegraphics[width=.9\textwidth]{figs/fig4.pdf}
 \caption{Surface (horizontal) KE (a) and SSH variance wavenumber spectra (b)
 in the 1/48$^\circ$ simulation. Solid lines
 are spectra based on hourly snapshots, dashed lines are spectra based on daily-averaged
 fields.}
 \label{fig3}
 \end{center}
\end{figure*}

@@

\section{Summary and Discussion}
Our main finding is that
upper-ocean divergent flows, dominated by inertia-gravity waves, undergo
a strong seasonal cycle that is out-of-phase with the seasonal cycle of
the quasi-balanced flows. Therefore, the present results suggest significant
seasonal modulation of the dominant upper-ocean dynamics at submesoscales
(10-100 km): quasi-balanced turbulence driven by shallow baroclinic instabilities
in late Winter/early Spring
and inertia-gravity waves dominate in late Summer/early Fall.

That the  surface velocity and SSH submesoscale variability may be
dominated by ageostrophic flows in
Summer/Fall has consequences for the interpretation of data from
the future SWOT and COMPIRA altimeter missions,
which will deliver SSH measurements at submesoscales. To the extent that
high-frequency flows are dominated by spatially incoherent internal tides and other
internal waves, it may be very difficult (if not impossible) to
separate SSH submesoscale variability associated with geostrophic motions
from high-frequency, ageostrophic flows. Thus, previous claims that
one will be able to easily obtain submesoscale surface
geostrophic velocities and monitor such seasonal cycle on global scales
\citep{sasaki_etal2014,qiu_etal2014} are overstated.

The present analysis focuses on a single patch of ocean in the
vicinities of the Kuroshio Extension, not
representative of other regions such as eastern boundary currents
and the middle of the subtropical gyre: care must be taken in overgeneralizing
these results.  The effects of smaller-scale/higher-frequency
``sub-submesoscale'' flows on the submesoscale surface velocity and SSH variability
are presently unknown.

>>>>

A key aspect of the LLC simulations is that they were forced by
the 16 most significant tidal components.
Because barotropic tides interact with topography and generate internal
tides that project on mesoscales to submesoscales
\citep[e.g., ][]{rocha_etal2016}, tidal forcing fundamentally distinguishes our analysis
from recent modeling studies that investigated upper-ocean dynamics
and its seasonality \citep{sasaki_etal2014,qiu_etal2014}. Details of the LLC simulations
are provided in the supplemental material.

@@

\begin{figure*}[ht]
\begin{center}
\hspace{-1.25cm}\includegraphics[width=.7\textwidth]{figs/fig1_1.pdf}\\
\vspace{-.125cm}
\includegraphics[width=.75\textwidth]{figs/fig1_2.pdf}
 \caption{(a) The study region with the subregion where the LLC outputs are
          analyzed. LLC 4320 (1/48$^\circ$) snapshots of surface vorticity (b and c) and a transect
          of potential density at 165$^\circ$E (d and e). The snapshots were
          taken at 00:00 UTC.}
\vspace{-1.5cm}
 \label{fig1}
 \end{center}
 \end{figure*}

@@

 \begin{figure*}[ht]
   \begin{center}
     \includegraphics[width=.75\textwidth]{figs/fig2.pdf}
  \caption{Time series of the root-mean-square (RMS) of vorticity (a) and
  lateral rate of strain (b), and horizontal divergence (c) in the LLC outputs and gridded AVISO data.}
  \label{fig2}
  \end{center}
\end{figure*}

@@

The seasonal cycle of the horizontal divergence, however, showcases the complexity
of the upper-ocean annual variability. If submesoscale eddies and fronts dominated
the near-surface variability all year, then the seasonal cycle of horizontal divergence,
vertical vorticity, and lateral strain rate would be in phase \citep[e.g.,][]{sasaki_etal2014}.
While there is a clear wintertime peak in divergence of daily-averaged velocity
(see dashed lines in figure \ref{fig2}a; RMS divergence $\sim$0.22 f in the 1/48$^\circ$
simulation), the hourly fields show
a stronger enhancement of
lateral divergence in late Summer/early Fall (RMS divergence $\sim$0.22 f the 1/48$^\circ$
simulation). Because the 1/48$^\circ$ simulation better resolve
smaller-scale submesoscale flows, the
winter peak in divergence is nearly as strong the summertime's. Of course, part of
these submesoscale fronts and eddies evolve relatively fast and there is no clear
temporal and spatial scale separation between those motions and inertia-gravity waves:
daily-averaging the velocity fields efficiently suppresses the summertime horizontally
divergent flows, but also reduces the wintertime lateral divergence by about 50$\%$.
As for the vorticity and strain rate, most of the lateral divergence is associated
with submesoscale flows: smoothing the the velocity fields with a 100km-cut-off
suppresses more than 80$\%$ of the RMS divergence.

\section{Seasonality in governing submesoscale dynamics}
The overall results of figure \ref{fig2}c strongly suggest that the
summertime submesoscale surface variability stems from different dynamics than
then wintertime variability. To better understand these differences, we calculate
vorticity-divergence and vorticity-strain joint probability distributions (jPDF).

The April vorticity-strain jPDF have a shape characteristic of submesoscale turbulence
 (see figures \ref{fig3}f-g). The alignment of vorticity and strain $\alpha \sim \zeta$
 with strong positive skewness are fingerprints of submesoscale fronts
 \citep{shcherbina_etal2013,mcwilliams2016}. The shape of the vorticity-strain
 jPDF is similar for hourly, daily-averaged, and residual velocities, although
 the vorticity skewness reduces 1.4 to 1.13 from hourly to daily-averaged.
 The April vorticity-divergence jPDFs, however, have a egg-like shape that which again highlight
 the strong skewness in vorticity. Thus wintertime submesoscale surface velocity
 is strongly dominated by submesoscale turbulence --- this is true even for
 superinertial submesoscale currents.

The October vorticity-strain jPDF show much weaker skewness (the vorticity skewness
is 0.68 and 0.67 for hourly and daily-averaged velocities). The shape of the
 vorticity-strain jPDF  appears to be a combination of two half-ellipses centered
 about $\zeta=0$, one with a 45$^\circ$ slope (characteristic of submesocale fronts
 that persist in Summer) and one with a very steep slope. The vorticity-divergence
 jPDF is significantly different than the April one. In particular, the jPDF
 for daily-averaged currents is still egg-shaped but hourly and residual are
 ellipses about $\zeta=0$, likely associated with inertia-gravity waves.

 Time series of vorticity and divergence PDFs (supplemental material) suggest
 a strong oscillation between these two regimes. In late Winter/early Spring
 the vorticity is strongly positively skewed, whereas the divergence is moderately
 negatively skewed as predicted by frontogenesis \citep{capet_etal2008a,mcwilliams2016}
. In late Summer/early Fall, the divergence is more stronger, but PDFs are much less skewed,
consistent with linear inertia-gravity waves.

\section{Projection onto horizontal scales}

To better understand the projection of these flows onto different horizontal
scales, we calculate wavenumber spectra of kinetic energy and sea-surface height
(SSH) variance.  Before calculating the spectra, time-mean and and spatial linear
trend were removed, and the ensuing fields were multiplied by a Hanning spectral ``window''.
We discuss only spectra for the 1/48$^\circ$, which simply extends the 1/24$^\circ$
towards smaller scales (supplemental material).

Figure \ref{fig3}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the spectra based on hourly
velocity snapshots are nearly indistinguishable within 95$\%$ confidence level
(compare solid lines in figure \ref{fig3}a). Consistent with the results of
\citet{rocha_etal2016} who analyzed a 5-month output of the LLC4320 simulation in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales smaller than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig3}a). But this suppression is more dramatic in October, when the horizontally
divergent flows peak. At scales
smaller than 100 km, 39$\%$ of the surface KE in April is accounted for by super-inertial
flows as opposed to 79$\%$ in October. The seasonality of subinertial submesoscale flows
is much more dramatic than the seasonal cycle of the total flow. This is consistent
with the results of \citet{sasaki_etal2014}, which are
based on daily-averaged velocity fields of a different model without tidal forcing
(P. Klein, personal communication).

These high-frequency submesoscale flows significantly project on the sea-surface.
In October, when the inertia-gravity waves are stronger near the surface and submesoscale
turbulence is weak, there is a dramatic difference between the spectra
based on hourly and daily-averaged SSH: at scales smaller than 100 km, the spectra
of hourly SSH roughly follows a $-2$ power-law, whereas the spectra of daily-averaged
SSH roughly follows a $-5$ power-law. At scales
smaller than 100 km, 33$\%$ of the SSH variance in April is accounted for by suprainertial
flows as opposed to 83$\%$ in October! The out-of-phase seasonal cycle of submesoscale
turbulence and near-surface inertia-gravity waves conspire to yield very weak seasonality
in the spectra of KE and SSH variance based on hourly field.

\section{Summary and Conclusion}
Our main finding is that
near-surface upper-ocean inertia-gravity waves, undergo
a strong seasonal cycle that is out-of-phase with the seasonal cycle of
the submesoscale turbulence. \cite{dasaro1978} showed that linear internal waves
velocity in the mixed-layer strongly depends on the density jump at the mixed-layer
base. In Summer, the shallow mixed-layer overlays a strong seasonal pycnocline,
and thus the internal waves projection onto the mixed-layer may be stronger
according to \cite{dasaro1978}'s argumets. An
alternative explanation is that the shape baroclinic modes experience significant
seasonal modulation, with significant summertime peak in their surface amplitude.
If the internal wave source has weak seasonal dependence, then a strong
seasonal cycle in the near-surface expression of internal waves is expected.
Internal tide generation is unlikely to vary significant seasonally, and therefore
it is plausible to expect a strong seasonality in the the near-surface expression of
 internal tides and
other small-scale internal waves generated through internal tide interactions.
Near-inertial wave generation peaks in Winter, but those waves project on large
horizontal scales. \textcolor{red}{(REFS)}.

Our work adds to recent studies that presented modeling \cite{sasaki_etal2014}
and observational \citep{callies_etal2015,buckingham_etal2016} evidence of vigorous seasonality in
submesoscale turbulence. We suggest that the dominance at inertia-gravity waves in
Summer is a consequence of both suppresion of submesoscale turbulence and
enhancement of inertia-gravity waves due to restratification of
the upper ocean. Of course, our study focuses on a single patch of ocean in the
vicinities of the Kuroshio Extension, which may be typical of mesoscale-rich
subtropical regions, but not
representative of other regions such as eastern boundary currents
and the middle of the subtropical gyre: care must be taken in overgeneralizing
these results.  We plan to report on the geographic variability of submesoscale
 seasonality  in a future study.

%The effects of smaller-scale/higher-frequency
%``sub-submesoscale'' flows on the submesoscale surface velocity and SSH variability
%are presently unknown.

% That the  surface velocity and SSH submesoscale variability may be
% dominated by ageostrophic flows in
% Summer/Fall has consequences for the interpretation of data from
% the future SWOT and COMPIRA altimeter missions,
% which will deliver SSH measurements at submesoscales. To the extent that
% high-frequency flows are dominated by spatially incoherent internal tides and other
% internal waves, it may be very difficult (if not impossible) to
% separate SSH submesoscale variability associated with geostrophic motions
% from high-frequency, ageostrophic flows. Thus, previous claims that
% one will be able to easily obtain submesoscale surface
% geostrophic velocities and monitor such seasonal cycle on global scales
% \citep{sasaki_etal2014,qiu_etal2014} are overstated.

\begin{figure*}[ht]
  \begin{center}
    \includegraphics[width=.75\textwidth]{figs/fig3.pdf}
 \caption{Seasonal variation of joint probability distributions: vorticity vs.
          divergence (a through e) and vorticity vs. strain rate (f through j).
          Dashed lines in (a) through (e) represent $\delta = \frac{\omega}{f_0}
          \zeta$ where $\omega = 2\pi/$(12 h, 3 h) and $f_0$ is the inertial frequency
          in the middle of the domain (32.5$^\circ$ N). Dashed lines in (f) through (j)
          represent pure strain flow $\alpha = \zeta$.}
 \label{fig3}
 \end{center}
\end{figure*}

\begin{figure*}[ht]
  \begin{center}
    \includegraphics[width=.9\textwidth]{figs/fig4.pdf}
 \caption{Surface (horizontal) KE (a) and SSH variance wavenumber spectra (b)
 in the 1/48$^\circ$ simulation. Solid lines
 are spectra based on hourly snapshots, dashed lines are spectra based on daily-averaged
 fields.}
 \label{fig4}
 \end{center}
\end{figure*}

@@

@@
